Uncovering underlying musical content in the temporal domain \\ \small Leveraging deep learning, inductive bias, and aural skills to learn embedding spaces for downstream MIR tasks

### 2023-03-12 figures/Figures CNN/Single_CNN.tex
Updates from Overleaf

\begin{SCfigure}[2\sidecaptionrelwidth][h]
	\centering
	\begin{tikzpicture}
		\node at (1.5,4){\begin{tabular}{c}input image\\or input feature map\end{tabular}};

@@

		\draw[fill=black,opacity=0.2,draw=black] (5.5,1.5) -- (7.5,1.5) -- (7.5,3.5) -- (5.5,3.5) -- (5.5,1.5);
		\draw[fill=black,opacity=0.2,draw=black] (5,1) -- (7,1) -- (7,3) -- (5,3) -- (5,1);
		\draw[fill=black,opacity=0.2,draw=black] (4.5,0.5) -- (6.5,0.5) -- (6.5,2.5) -- (4.5,2.5) -- (4.5,0.5);
		\draw[fill=black,opacity=0.2,draw=black] (4,0) -- (6,0) -- (6,2) -- (4,2) -- (4,0);
	\end{tikzpicture}
	\caption[Illustration of a convolutional layer.]{Illustration of a single convolutional layer. If layer $l$ is a convolutional layer, the input image (if $l = 1$) or a feature map of the previous layer is convolved by different filters to yield the output feature maps of layer $l$.}
	\label{fig:convolutional-layer}
\end{SCfigure}

>>>>

\begin{SCfigure}[2\sidecaptionrelwidth][h]
	\centering
	\begin{tikzpicture}
		\node at (1.5,4){\begin{tabular}{c}input data\\or input feature map\end{tabular}};

@@

		\draw[fill=black,opacity=0.2,draw=black] (5.5,1.5) -- (7.5,1.5) -- (7.5,3.5) -- (5.5,3.5) -- (5.5,1.5);
		\draw[fill=black,opacity=0.2,draw=black] (5,1) -- (7,1) -- (7,3) -- (5,3) -- (5,1);
		\draw[fill=black,opacity=0.2,draw=black] (4.5,0.5) -- (6.5,0.5) -- (6.5,2.5) -- (4.5,2.5) -- (4.5,0.5);
		\draw[fill=black,opacity=0.2,draw=black] (4,0) -- (6,0) -- (6,2) -- (4,2) -- (4,0);
	\end{tikzpicture}
	\caption[Illustration of a convolutional layer.]{Illustration of a single convolutional layer. If layer $l$ is a convolutional layer, the input data (if $l = 1$) or a feature map of the previous layer is convolved by different filters to yield the output feature maps of layer $l$.}
	\label{fig:convolutional-layer}
\end{SCfigure}

### 2023-03-12 introduction/Subsections/Introduction.tex
Updates from Overleaf

The purpose of this work is to generalize the fundamental harmonic structure of music to facilitate its application in downstream tasks. Music, being a complex and fascinating art form, has been extensively studied by scholars and musicians for centuries. Understanding how music is organized and composed is critical, and music theory plays a vital role in this regard. Analyzing music structure poses a significant challenge as it comprises several elements, including melody, harmony, rhythm, and form.

In recent years, the utilization of neural networks has made it possible to create low-dimensional musical embeddings that capture crucial musical features and enable the computation of semantic similarity between musical passages. By acquiring an embedding function $f(x) \in \mathbb{R}^d$ that maps a given passage $x$ to a vector in the real space $\mathbb{R}^d$, researchers can measure the semantic similarity between two musical passages using the Euclidean distance, as represented by the formula $D_{i,j} = ||f(x_i) - f(x_j)||^2$. The effectiveness of such systems depends on the embedding space's ability to interpret music in a way that aligns with human perception. Therefore, a suitable embedding space is crucial to ensure that the output aligns with human expectations.

The thesis under discussion explores the various musical elements that constitute music structure analysis, which includes melody, harmony, rhythm, and form. Aural skills are required to comprehend musical works through listening, as they involve perceptual and cognitive abilities. The underlying musical structure refers to the systematic arrangement of musical elements that results in a coherent and expressive musical work.

To provide a comprehensive analysis of musical works from diverse styles and genres, the thesis employs computational techniques from MIR and incorporates a musicological perspective to contextualize the findings. The objective is to offer fresh insights into these components of musical composition and advance our understanding of their relationships.

The study investigates the correlation between MSA, aural skills, and the underlying musical structure in musical composition. Through this research, the thesis aims to provide a better understanding of musical composition components, their relationships, and their practical applications.

>>>>

Music is an essential part of human culture, history, and society. It has existed in our lives since ancient times and has evolved, reflecting changes in our cultures, traditions, and beliefs. Despite its importance, the study of music has always been a challenging and complex task due to its subjective nature and the vast array of musical styles, genres, and cultures that exist worldwide.

Humanity has been studying music to understand its underlying structure for millennia. From the ancient Greeks, who believed that music had a mathematical and cosmic harmony, to the early works of theorists such as Rameau, Fux, and Schenker, and more recent approaches such as generative theory and corpus studies, music has been a subject of fascination and inquiry. This rich music theory and analysis history provides a valuable foundation for guiding and evaluating MIR algorithms and models.

We all humbly seek to unlock the secrets of music's complex and multifaceted nature.

Music Information Retrieval (MIR) is a research area that has emerged in the last decades as an interdisciplinary field combining engineering, musicology, and neuroscience. MIR aims to develop algorithms and techniques to automatically analyze, classify, and retrieve musical information from audio signals. In my opinion, MIR research has been traditionally focused on technical aspects, such as digital signal processing (DSP), machine learning, and pattern recognition, neglecting other relevant domains, such as perception, cognition, music theory, aural skills, and musicology.

The reason behind this is non-other but the high implementation complexity for such heuristics and abstract fields of study.

Let's pick a core MSA task as a sample to display such complexity: Music Structure Analysis (MSA) is an interdisciplinary field that aims to understand the structure of music. However, due to subjectivity, ambiguity, and data scarcity, audio-based MSA faces challenges such as boundary placement ambiguity and similarity quantification. The main principles of MSA were initially defined as homogeneity, novelty, and repetition, with the addition of regularity.

The checkerboard kernel technique is a simple and effective method for Music Structure Analysis (MSA) based on the homogeneity principle. The kernel, with a checkerboard-like structure, is convolved over the main diagonal of a Self-Similarity Matrix (SSM) such as $S_{ij}$ is the similarity between time frames $i$ and $j$, $\vec{x}_i$ is the vector representation of time frame $i$, and $\left| \vec{x}_i \right|$ is the Euclidean norm of vector $\vec{x}_i$. The dot product of two vectors $\vec{x}_i$ and $\vec{x}_j$ is denoted by $\vec{x}_i \cdot \vec{x}_j$.

\input{introduction/subsections/equations/SSM equation}

This yields a novelty curve that highlights sudden changes in the selected musical features (e.g. chroma, MFCCs, beat-synchronous features, onset and offset detection, and harmonic and percussive separation) from which to extract the segment boundaries.

\input{introduction/subsections/equations/Novelty equation}

Mathematical operations are used on predefined principles such as homogeneity and repetition to analyze a musical structure. It relies on analyzing features extracted from a musical signal and identifying peaks in a novelty curve to extract segment boundaries. In contrast, human aural skills rely on the perception and interpretation of music by human listeners based on cultural and academical background and can capture subtle nuances of musical structure, but those still are subjective and can vary across listeners.

Combining technical tasks, aural skills, and music perception is challenging due to the complexity and subjectivity of musical perception, discrepancies between the features extracted by DSP techniques and the features perceived by humans, and the variability of musical signals and individual differences in musical perception. DSP techniques may not always reflect the way humans perceive music, and may not capture the subtle variations in timing or harmonic relationships that are critical to musical perception. Variability in musical signals due to instrumentation, genre, and performance style further complicates the development of DSP techniques that can be generalized to different musical contexts.

### 2023-03-12 introduction/Subsections/Objectives.tex
Updates from Overleaf

As a professional music transcriber, I aim to understand how deep neural networks (DNNs) filter musical features, listen to the NN layers from a musician's perspective and analyze the resulting data. I aim to approach the technology from a technical and musical standpoint to understand how DNNs process and extract musical information. By doing so, I aim to enhance the accuracy and efficiency of the process and contribute to developing AI technology for music analysis. By combining my expertise in music with my -not fully developed- knowledge of DNNs, I hope to bring a unique perspective to the field and contribute to the advancement of music analysis.

@@

>>>>

The purpose of this work is to generalize the fundamental harmonic structure of music to facilitate its application in downstream music information retrieval (MIR) tasks. Music, being a complex and fascinating art form, has been extensively studied by scholars and musicians for centuries. Understanding how music is organized and composed is critical, and music theory plays a vital role in this regard. Analyzing music structure poses a significant challenge as it comprises several elements, including melody, harmony, rhythm, and form.

In recent years, the utilization of neural networks has made it possible to create low-dimensional musical embeddings that capture crucial musical features and enable the computation of semantic similarity between musical passages. By acquiring an embedding function $f(x) \in \mathbb{R}^d$ that maps a given passage $x$ to a vector in the real space $\mathbb{R}^d$, researchers can measure the semantic similarity between two musical passages using the Euclidean distance, as represented by the formula $D_{i,j} = ||f(x_i) - f(x_j)||^2$. The effectiveness of such systems depends on the embedding space's ability to interpret music in a way that aligns with human perception. Therefore, a suitable embedding space is crucial to ensure that the output aligns with human expectations.

The thesis under discussion explores the various musical elements that constitute music structure analysis, which includes melody, harmony, rhythm, and form. Aural skills are required to comprehend musical works through listening, as they involve perceptual and cognitive abilities. The underlying musical structure refers to the systematic arrangement of musical elements that results in a coherent and expressive musical work.

To provide a comprehensive analysis of musical works from diverse styles and genres, the thesis employs computational techniques from MIR and incorporates a musicological perspective to contextualize the findings. The objective is to offer fresh insights into these components of musical composition and advance our understanding of their relationships.

@@

We will argue that MIR research should shift from a purely technical-centered approach to a more balanced and comprehensive one that incorporates other relevant domains.

The study investigates the correlation between XXX, aural skills, and the underlying musical structure in musical composition. Through this research, the thesis aims to provide a better understanding of musical composition components, their relationships, and their practical applications.

### 2023-03-12 introduction/subsections/equations/Novelty equation.tex
Updates from Overleaf

Where:
\begin{itemize}
\item $N$ is the number of time frames.
\item $S_{ij}$ is the similarity between time frames $i$ and $j$.
\item $S_{ii}$ is the similarity between time frame $i$ and itself.
\item $n_i$ is the novelty value for time frame $i$.
\end{itemize}

The numerator of the expression represents the sum of similarities between time frame $i$ and all other time frames, minus the average similarity across all time frames. The denominator of the expression is the standard deviation of the similarities, which is used to normalize the values to ensure that the novelty values are between -1 and 1.

>>>>

Where $N$ is the number of time frames, $S_{ij}$ is the similarity between time frames $i$ and $j$, $S_{ii}$ is the similarity between time frame $i$ and itself, and $n_i$ is the novelty value for time frame $i$.

The numerator of the expression represents the sum of similarities between time frame $i$ and all other time frames minus the average similarity across all time frames. The denominator of the expression is the standard deviation of the similarities, which is used to normalize the values to ensure that the novelty values are between -1 and 1.
